---
title: "Analysis of PM2.5 in the U.S. and its Impact on Public Health"
author: "Jeffrey Hunter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    keep_md: no
    df_print: paged
  pdf_document:
    toc: yes
    df_print: kable
    number_sections: false
    fig_caption: yes
    highlight: tango
    dev: pdf
  word_document:
    toc: yes
    df_print: paged
    keep_md: no
---

## Course Project

**Exploratory Data Analysis Course Project 2**

Peer-graded Assignment

* This course project is available on GitHub

    <a href="https://github.com/oraclejavanet/exploratory-data-analysis-course-project-2" target="_top">Analysis of PM2.5 in the United States and its Impact on Public Health</a>

## Synopsis

Fine particulate matter (PM2.5) is an ambient air pollutant for which there is
strong evidence that it is harmful to human health. In the United States, the
Environmental Protection Agency (EPA) is tasked with setting national ambient
air quality standards for fine PM and for tracking the emissions of this
pollutant into the atmosphere. Approximately every 3 years, the EPA releases its
database on emissions of PM2.5. This database is known as the National Emissions
Inventory (NEI). You can read more information about the NEI at the
[EPA National Emissions Inventory web site](http://www.epa.gov/ttn/chief/eiinformation.html).

For each year and for each type of PM source, the NEI records how many tons of
PM2.5 were emitted from that source over the course of the entire year. The data
that you will use for this assignment are for 1999, 2002, 2005, and 2008.

## Assignment

The overall goal of this assignment is to explore the National Emissions
Inventory database and see what it says about fine particulate matter pollution
in the United states over the 10-year period 1999–2008. You may use any R
package you want to support your analysis.

The deliverable for this assignment is to address the questions and tasks
provided in the [Questions](#questions) section of this report.

## Environment Setup

```{r setup, include = FALSE}
# set knitr options
knitr::opts_chunk$set(echo = TRUE, fig.path='figures/')

# clear global workspace (including hidden objects) for this session
rm(list = ls(all.names = TRUE))

# free up memory and display statistics on free memory
gc()

# disable scientific notation for numbers
options(scipen = 1)
```

Load packages used in this analysis.

```{r load-packages, echo = TRUE}
if (!require(ggplot2)) {
    install.packages("ggplot2")
    library(ggplot2, warn.conflicts = FALSE)
}
if (!require(dplyr)) {
    install.packages("dplyr")
    library(dplyr, warn.conflicts = FALSE)
}
if (!require(scales)) {
    install.packages("scales")
    library(scales, warn.conflicts = FALSE)
}
if (!require(stringi)) {
    install.packages("stringi")
    library(stringi, warn.conflicts = FALSE)
}
```

Display session information.

```{r display-session-info, echo = TRUE}
sessionInfo()
```

## Load Data

Download the compressed data file from the source URL (if not found locally),
uncompress it and then load the two serialized R objects via the `readRDS`
function. Prior to processing the data, validate the downloaded data file and
loaded dataset by checking the file size and dimensions respectively.

```{r load-data, echo = TRUE, cache = TRUE}
setwd("~/repos/coursera/github-assignments/exploratory-data-analysis-course-project-2")
extNeiDataFileURL <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"
extNeiDataFile <- "data/exdata_data_NEI_data.zip"
neiDataFile <- "data/summarySCC_PM25.rds"
sccDataFile <- "data/Source_Classification_Code.rds"
if (!file.exists('data')) {
    dir.create('data')
}
if (!file.exists(extNeiDataFile)) {
    download.file(url = extNeiDataFileURL, destfile = extNeiDataFile)
    unzip(extNeiDataFile, exdir = "data")
}
stopifnot(file.size(extNeiDataFile) == 30643310)
NEI <- readRDS(neiDataFile)
SCC <- readRDS(sccDataFile)
stopifnot(dim(NEI) == c(6497651, 6))
stopifnot(dim(SCC) == c(11717, 15))
```

## Questions

1. Have total emissions from PM2.5 decreased in the United States from 1999 to
   2008? Using the **base** plotting system, make a plot showing the *total*
   PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.

   **Answer**
   
   The answer to the question is yes. The plot below shows that total emissions
   from PM2.5 has decreased in the United States from 1999 to 2008.
   
    ```{r question-1, echo = TRUE}
    # summarize total emissions by year for U.S.
    totalByYear <- NEI %>%
                   group_by(year) %>%
                   filter(year == 1999|2002|2005|2008) %>%
                   summarize(Total.Emissions = sum(Emissions))
    
    # manually specify axis parameters
    xYears <- c(1999, 2002, 2005, 2008)
    yEmiss <- pretty(totalByYear$Total.Emissions/10^3, n = 4)
    
    # create plot with custom axes
    plot(totalByYear$year,
         totalByYear$Total.Emissions/10^3,
         type = "b",
         bty = "l",
         lwd = 3,
         pch = 19,
         col = rgb(0.2, 0.4, 0.6, 0.8),
         axes = FALSE,
         yaxt = "n",
         xlab = "Year",
         ylab = "Total PM2.5 Emissions (in kilotons)",
         main = "Total PM2.5 Emissions in the U.S. 1999-2008")
    axis(1, at = xYears, labels = xYears)
    axis(2, at = yEmiss, labels = prettyNum(yEmiss, big.mark = ","))
    ```

1. Have total emissions from PM2.5 decreased in the **Baltimore City**, Maryland
   (`fips == "24510"`) from 1999 to 2008? Use the **base** plotting system to
   make a plot answering this question.

   **Answer**
   
   The answer to the question is yes. The plot below shows that total emissions
   from PM2.5 has decreased in Baltimore City, MD from 1999 to 2008. The plot
   also shows that the decrease in emissions from 1999 to 2008 did not follow
   a consistent negative trend. A sharp rise in emissions was observed between
   2002 and 2005.
   
    ```{r question-2, echo = TRUE}
    # summarize total emissions by year for Baltimore (fips == "24510")
    totalBaltimore <- NEI %>%
                      filter(fips == "24510") %>%
                      group_by(year) %>%
                      filter(year == 1999|2002|2005|2008) %>%
                      summarize(Total.Emissions = sum(Emissions))
    
    # manually specify axis parameters
    xYears <- c(1999, 2002, 2005, 2008)
    yEmiss <- pretty(totalBaltimore$Total.Emissions/10^3, n = 4)
    
    # create plot with custom axes
    plot(totalBaltimore$year,
         totalBaltimore$Total.Emissions/10^3,
         type = "b",
         bty = "l",
         lwd = 3,
         pch = 19,
         col = rgb(0.2, 0.4, 0.6, 0.8),
         axes = FALSE,
         xlab = "Year",
         ylab = "Total PM2.5 Emissions (in kilotons)",
         main = "Total PM2.5 Emissions in Baltimore City, MD 1999-2008")
    axis(1, at = xYears, labels = xYears)
    axis(2, at = yEmiss, labels = yEmiss)
    ```

1. Of the four types of sources indicated by the `type` (point, nonpoint,
   onroad, nonroad) variable, which of these four sources have seen decreases
   in emissions from 1999–2008 for **Baltimore City**? Which have seen increases
   in emissions from 1999–2008? Use the **ggplot2** plotting system to make a
   plot answer this question.

   **Answer**
   
   The plot below shows that total emissions from PM2.5 has decreased in
   Baltimore City, MD from 1999 to 2008 for the *nonpoint*, *on-road* and
   *non-road* source types whereas the *point* source type has seen an increase
   in emissions during the same time period.
   
    ```{r question-3, echo = TRUE}
    # summarize total emissions by source type and year for Baltimore (fips == "24510")
    totalBaltimoreBySource <- NEI %>%
                              filter(fips == "24510") %>%
                              group_by(type, year) %>%
                              filter(year == 1999|2002|2005|2008) %>%
                              summarize(Total.Emissions = sum(Emissions))
    
    # optional: specify the order of the plots
    totalBaltimoreBySource$type <- factor(totalBaltimoreBySource$type,
                                          levels = c("POINT", "NONPOINT", "ON-ROAD", "NON-ROAD"))

    # create plot 
    gTotBalSrc <- ggplot(data = totalBaltimoreBySource, aes(x = factor(year),
                                                            y = Total.Emissions,
                                                            fill = type)) +
         geom_bar(stat = "identity") +
         facet_grid(. ~ type) +
         xlab("Year") + 
         ylab("Total PM2.5 Emissions (in tons)") +
         guides(fill = FALSE) +
         theme(plot.title = element_text(size = 14, hjust = 0.5, vjust = 0.5),
               axis.text.x = element_text(angle = 45,
                                          hjust = 0.5,
                                          vjust = 0.5,
                                          margin = margin(b = 10))) +
         scale_y_continuous(labels = comma) +
         ggtitle("Total PM2.5 Emissions in\nBaltimore City, MD by Source Type 1999-2008")
    print(gTotBalSrc)
    ```

1. Across the United States, how have emissions from coal combustion-related
   sources changed from 1999–2008?

   **Answer**

   The plot below shows that emissions from coal combustion-related sources
   has decreased from 1999-2008.
   
    ```{r question-4, echo = TRUE}
    # select SCC values in SCC where Short.Name includes "[Cc]oal"
    coalSCC <- subset(SCC,
                      stri_detect_regex(Short.Name, "Coal", case_insensitive = TRUE),
                      select = c(SCC))
    
    # convert to numeric factor
    coalSCC <- coalSCC$SCC
    
    # select year, Emissions from NEI where SCC in (filtered list of SCC values) 
    coalNEI <- subset(NEI, SCC %in% coalSCC, select = c(Emissions, year))
    
    # summarize total coal emissions by year for U.S.
    totalCoalNEI <- coalNEI %>%
                    group_by(year) %>%
                    filter(year == 1999|2002|2005|2008) %>%
                    summarize(Total.Emissions = sum(Emissions))

    # manually specify axis parameters
    xYears <- c(1999, 2002, 2005, 2008)
    yEmiss <- pretty(totalCoalNEI$Total.Emissions/10^3, n = 4)
    
    # create plot with custom axes
    plot(totalCoalNEI$year,
         totalCoalNEI$Total.Emissions/10^3,
         type = "b",
         bty = "l",
         lwd = 3,
         pch = 19,
         col = rgb(0.2, 0.4, 0.6, 0.8),
         axes = FALSE,
         xlab = "Year",
         ylab = "Total PM2.5 Emissions (in kilotons)",
         main = "Total PM2.5 Emissions from\nCoal Combustion-Related Sources in the U.S. 1999-2008")
    axis(1, at = xYears, labels = xYears)
    axis(2, at = yEmiss, labels = yEmiss)
    ```

1. How have emissions from motor vehicle sources changed from 1999–2008 in
   **Baltimore City**?

   **Answer**

   The plot below shows that emissions from motor vehicle sources has decreased
   from 1999-2008.

    ```{r question-5, echo = TRUE}
    # select SCC values in SCC where EI.Sector includes "[Vv]ehicle"
    vehicleSCC <- subset(SCC,
                         stri_detect_regex(EI.Sector, "Vehicle", case_insensitive = TRUE),
                         select = c(SCC))
    
    # convert to numeric factor
    vehicleSCC <- vehicleSCC$SCC
    
    # select year, Emissions, fips from NEI where SCC in (filtered list of SCC values) 
    vehicleNEI <- subset(NEI, SCC %in% vehicleSCC, select = c(Emissions, year, fips))
    
    # summarize vehicle emissions by year for Baltimore (fips == "24510")
    totalVehicleNEI <- vehicleNEI %>%
                       filter(fips == "24510") %>%
                       group_by(year) %>%
                       filter(year == 1999|2002|2005|2008) %>%
                       summarize(Total.Emissions = sum(Emissions))

    # manually specify axis parameters
    xYears <- c(1999, 2002, 2005, 2008)
    yEmiss <- pretty(totalVehicleNEI$Total.Emissions, n = 4)
    
    # create plot with custom axes
    plot(totalVehicleNEI$year,
         totalVehicleNEI$Total.Emissions,
         type = "b",
         bty = "l",
         lwd = 3,
         pch = 19,
         col = rgb(0.2, 0.4, 0.6, 0.8),
         axes = FALSE,
         xlab = "Year",
         ylab = "Total PM2.5 Emissions (in tons)",
         main = "Total PM2.5 Emissions in Baltimore City, MD for\nMotor Vehicles 1999-2008")
    axis(1, at = xYears, labels = xYears)
    axis(2, at = yEmiss, labels = yEmiss)
    ```
   
1. Compare emissions from motor vehicle sources in **Baltimore City** with
   emissions from motor vehicle sources in **Los Angeles County**, California
   (`fips == "06037"`). Which city has seen greater changes over time in motor
   vehicle emissions?

   **Answer**

   The plot below shows that emissions from motor vehicle sources in
   Baltimore City has decreased from 1999-2008 whereas Los Angeles County
   has seen an increase from motor vehicle sources during the same time period.

    ```{r question-6, echo = TRUE}
    # select SCC values in SCC where EI.Sector includes "[Vv]ehicle"
    vehicleSCC <- subset(SCC,
                         stri_detect_regex(EI.Sector, "Vehicle", case_insensitive = TRUE),
                         select = c(SCC))
    
    # convert to numeric factor
    vehicleSCC <- vehicleSCC$SCC
    
    # select year, Emissions, fips from NEI where SCC in (filtered list of SCC values) 
    vehicleNEI <- subset(NEI, SCC %in% vehicleSCC, select = c(Emissions, year, fips))
    
    # summarize vehicle emissions by year for Baltimore (fips == "24510")
    vehicleBaltimoreNEI <- vehicleNEI %>%
                           filter(fips == "24510") %>%
                           mutate(city = "Baltimore") %>%
                           group_by(city, year) %>%
                           filter(year == 1999|2002|2005|2008) %>%
                           summarize(Total.Emissions = sum(Emissions))
    
    # summarize vehicle emissions by year for Los Angeles (fips == "06037")
    vehicleLosAngelesNEI <- vehicleNEI %>%
                            filter(fips == "06037") %>%
                            mutate(city = "Los Angeles") %>%
                            group_by(city, year) %>%
                            filter(year == 1999|2002|2005|2008) %>%
                            summarize(Total.Emissions = sum(Emissions))

    # combine both datasets
    totalVehicleNEI <- rbind(vehicleBaltimoreNEI, vehicleLosAngelesNEI)

    # optional: specify the order of the plots
    totalVehicleNEI$city <- factor(totalVehicleNEI$city, levels = c("Baltimore", "Los Angeles"))
    
    # create plot 
    gtotalVehicleNEI <- ggplot(data = totalVehicleNEI, aes(x = factor(year),
                                                           y = Total.Emissions,
                                                           fill = city)) +
         geom_bar(stat = "identity") +
         facet_grid(. ~ city) +
         xlab("Year") + 
         ylab("Total PM2.5 Emissions (in tons)") +
         guides(fill = FALSE) +
         theme(plot.title = element_text(size = 14, hjust = 0.5, vjust = 0.5),
               axis.text.x = element_text(angle = 45,
                                          hjust = 0.5,
                                          vjust = 0.5,
                                          margin = margin(b = 10))) +
         scale_y_continuous(labels = comma) +
         ggtitle("Total PM2.5 Emissions\nBaltimore City, MD versus Los Angeles, CA\nfor Motor Vehicles 1999-2008")
    print(gtotalVehicleNEI)
    ```
